---
title: "Social Media & Mental Health"
author: "Cixian Huang, Izayah Qiu"
date: today
format:
  dashboard:
    theme: cosmo
    embed-resources: true
---

```{r}
library(tidyverse)
library(plotly)
```

```{r}
df1 <- read.csv("test.csv", header = TRUE)
df2 <- read.csv("mental_health2425v8in.csv", header = TRUE)
df2 <- df2 |>
  mutate(Age = 2026 - year_1) |>
  rename(Gender = sex)

df1 <- df1 |>
  mutate(Age = as.numeric(Age)) |>
  distinct()

combined_data <- left_join(df1, df2, by = c("Age", "Gender"))
combined_data <- combined_data |>
  mutate(
    Total_Mental_Health = phq9_COMP + gad7_COMP,
    Usage_Level = ifelse(
      Daily_Usage_Time..minutes. >= median(Daily_Usage_Time..minutes., na.rm = TRUE),
      "High Usage",
      "Low Usage"
    )
  ) |>
  filter(
    !is.na(Daily_Usage_Time..minutes.),
    !is.na(phq9_COMP),
    !is.na(gad7_COMP)
  )
correlation_value <- cor(
  combined_data$Daily_Usage_Time..minutes.,
  combined_data$Total_Mental_Health
)

cor_rounded <- round(correlation_value, 3)

n_participants <- nrow(combined_data)

avg_usage <- round(mean(combined_data$Daily_Usage_Time..minutes., na.rm = TRUE), 1)


avg_mh <- round(mean(combined_data$Total_Mental_Health, na.rm = TRUE), 1)
```

# Overview

#### About This Project

In the age of technology, social media is one of the most dominant forms of communication and entertainment. Platforms such as Instagram, TikTok, and X allow people to share experiences, interact with others, and consume large amounts of content in a short period of time.

While these platforms bring many positive benefits — such as connecting with others and raising awareness — researchers and public health experts have also raised concerns about their impact on mental health. According to the **American Psychological Association**, excessive social media use has been linked to increased anxiety, depression, and feelings of loneliness, particularly among younger users.

> *Teens "find it difficult to picture their lives without social media."* — PubMed Central

Our project explores how different patterns of social media use may relate to emotional states and mental well-being. Rather than looking only at how long someone spends online, we analyze several forms of engagement including posting, messaging, and receiving likes or comments.

**Sources:**  
[APA — Protecting Teens on Social Media](https://www.apa.org/monitor/2023/09/protecting-teens-on-social-media)  
[PubMed Central — Social Media & Mental Health](https://pmc.ncbi.nlm.nih.gov/articles/PMC10476631/)

### Column {width=45%}

#### Research Questions

---

**Question 1**  
Using PHQ-9 & GAD-7 scores, is higher daily social media usage associated with worse overall mental health outcomes?

---

**Question 2**  
Do different types of social media engagement correspond with different emotional outcomes?

---

**Question 3**  
Are there demographic differences — such as age or gender — in how social media use relates to emotional well-being?

---

#### Data Sources

| Dataset | Source | Observations |
|--------|--------|-------------|
| Social Media Usage & Emotional Well-Being | [Kaggle](https://www.kaggle.com/datasets/emirhanai/social-media-usage-and-emotional-well-being) | 103 |
| Social Media & Mental Health | [UMD BSOS](https://bsos-data.umd.edu/dataset/social-media-and-mental-health) | Multiple |

> Both datasets rely on self-reported data. Results should be interpreted as **exploratory**, not causal.

# Analysis

## Row

### Column {.tabset width=60%}

#### Q1: Usage vs Mental Health

```{r}
#| fig-alt: "Scatterplot of daily usage versus mental health score, colored by age, with a positive trend line."
p1 <- ggplot(
    combined_data,
    aes(
      x    = Daily_Usage_Time..minutes.,
      y    = Total_Mental_Health,
      color = Age,
      text = paste0(
        "Age: ", Age,
        "<br>Daily Usage: ", Daily_Usage_Time..minutes., " min",
        "<br>PHQ-9 + GAD-7: ", Total_Mental_Health
      )
    )
  ) +
  geom_point(alpha = 0.65, size = 2.5) +
  geom_smooth(method = "lm", se = FALSE, color = "tomato", linewidth = 1) +
  scale_color_viridis_c(option = "plasma") +
  labs(
    title   = "Daily Social Media Usage vs. Mental Health Score",
    x       = "Daily Usage (Minutes)",
    y       = "PHQ-9 + GAD-7 Score",
    color   = "Age",
    caption = paste0("Pearson r = ", cor_rounded,
                     " — a moderate positive association")
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
ggplotly(p1, tooltip = "text")
```

#### Q2: Emotion & Usage

```{r}
#| fig-alt: "Boxplot of daily usage time across dominant emotion categories, ordered by median."
p2 <- ggplot(
    combined_data,
    aes(
      x    = reorder(Dominant_Emotion, Daily_Usage_Time..minutes., median),
      y    = Daily_Usage_Time..minutes.,
      fill = Dominant_Emotion
    )
  ) +
  geom_boxplot(outlier.shape = 21, outlier.size = 2, alpha = 0.8) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title    = "Daily Usage Time by Dominant Emotion",
    x        = "Dominant Emotion",
    y        = "Daily Usage (Minutes)",
    caption  = "Emotions are ordered by median daily usage time."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    plot.title      = element_text(face = "bold"),
    axis.text.x     = element_text(angle = 20, hjust = 1)
  )

ggplotly(p2, tooltip = c("y", "fill"))
```

#### Q3: High vs Low Usage

```{r}
#| fig-alt: "Boxplot comparing mental health scores between high and low social media usage groups."
p3 <- ggplot(
    combined_data,
    aes(
      x    = Usage_Level,
      y    = Total_Mental_Health,
      fill = Usage_Level
    )
  ) +
  geom_boxplot(width = 0.5, outlier.shape = 21, outlier.size = 2, alpha = 0.85) +
  scale_fill_manual(values = c("High Usage" = "#E07B54", "Low Usage" = "#5B8DB8")) +
  labs(
    title   = "Mental Health Score by Usage Level",
    x       = "Usage Level (relative to median)",
    y       = "Total PHQ-9 + GAD-7 Score",
    caption = "High = at or above median daily usage. Low = below median."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    plot.title      = element_text(face = "bold")
  )

ggplotly(p3, tooltip = c("y", "fill"))
```

### Column {width=40%}

#### Key Takeaways

---

**Question 1 — Usage & Mental Health**

The scatterplot shows a moderate positive trend (r = `r cor_rounded`): participants who spend more time on social media daily tend to report higher combined PHQ-9 + GAD-7 scores. Most participants cluster between **50–120 minutes/day** with scores of **10–25**.

---

**Question 2 — Emotion & Usage**

Daily usage time varies noticeably across dominant emotion groups. Users who report positive emotions (e.g., *happy*, *relaxed*) tend to spend more time on social media, while those reporting *bored* or *sad* show a wider spread of usage patterns.

---

**Question 3 — High vs Low Users**

High-usage participants show slightly higher total mental health burden than low-usage participants, but the difference is modest. The overlap between groups suggests usage alone is not a strong predictor of mental health outcomes.

---

> These findings are **associational** — not causal. Many other factors influence mental health.

# Conclusion

## Row

### Column {width=60%}

#### Summary of Findings

This project explored whether social media usage patterns are related to mental health indicators across two public datasets.

**What we found:**

- There is a **moderate positive correlation** (r = `r cor_rounded`) between daily social media usage time and combined PHQ-9 + GAD-7 scores.
- Users reporting **positive dominant emotions** (happy, relaxed) tended to use social media more, while those reporting negative emotions showed more variable usage.
- **High-usage participants** had slightly higher mental health burden scores than low-usage participants, though the groups overlapped considerably.

**Limitations to keep in mind:**

- Both datasets rely on **self-reported data**, which can introduce recall or social desirability bias.
- The first dataset has only **103 observations**, limiting generalizability.
- Joining by age and gender alone may introduce **mismatches** between respondents across datasets.
- The analysis is **cross-sectional** — we cannot determine whether usage causes worse mental health or vice versa.

### Column {width=40%}

#### Future Directions

Expanding this research could involve:

**Larger, more representative samples** to improve statistical power and generalizability

**Longitudinal data** to track how mental health changes alongside social media habits over time

**Platform-specific analysis** — Instagram, TikTok, and X may have distinct effects

**Advanced modeling** (e.g., regression controlling for demographics, sleep, and exercise) to isolate the effect of social media

**Subgroup analyses** by age group or gender to better understand demographic differences

---

> *"The relationship between social media and mental health is complex — it's not simply good or bad, but depends on how, why, and how much people use it."*
